---
title: "Ptychodera Cisreg Development: Re-analysis of Branchiostoma floridae"
author: "Alberto Perez-Posada"
date: "6/21/2023"
output: html_document
---

```{r setup, include=FALSE}
dir <- '/home/ska/aperpos/projects/ptychodera_cisreg_development/'
fcha <- function(){ gsub("-","",Sys.Date()) }

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = dir)
options(scipen=999)
Sys.setenv(VROOM_CONNECTION_SIZE=5000072)
```

## About

In this markdown, we will do a re-analysis of the transcriptome of *Branchiostoma floridae*.

## Load libraries

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(stringr)
library(BiocGenerics)
library(EDASeq)
library(DESeq2)
library(tximport)
library(limma)
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
```

## Load functions

```{r load_functions, message = FALSE}
source("code/r_code/r_functions/sourcefolder.R")

sourceFolder(
  "code/r_code/r_functions",
  recursive = TRUE
  )

sourceFolder(
  "code/r_code/r_general",
  recursive = TRUE
  )
```

## Load data

We will load a sample table with all the information of each developmental transcriptome, in a similar way we did with Ptychodera.

```{r}
samplesdir <- 
  "outputs/comparative/reanalysis/bflo/kallisto/abundances/"

sampleTable <- 
  read.table(
    "outputs/comparative/reanalysis/bflo/kallisto/abundances/sample_table.tsv",
    header=F
  )

sampleTable <- 
  merge(
    sampleTable,
    as.data.frame(c(list.dirs(samplesdir,full.names = F)[-1])),
    by=1
    )

colnames(sampleTable) <- c("sample","condition","batch")

sampleTable$batch <- 
  factor(sampleTable$batch, levels = unique(sort(sampleTable$batch)))

sampleTable$condition <- factor(sampleTable$condition)

rownames(sampleTable) <- sampleTable[,1]
sampleTable$sample <- NULL

bflo_condition_x <- sampleTable$condition
```

We will also load a dictionary between the gene ids of our mapping and the gene ids of the proteome used in comparative analyses (OMA, Orthofinder). This will be useful later in the future.

```{r load_dict}
bf_bflo <- read.table("outputs/comparative/oma/dicts/BRAFL.dct", col.names = c("key","value"))
```

## Tximport

We load the transcriptome mapping from kallisto here using the data of the sample table:

```{r tximport}
library(tximport)
files <- 
  list.files(samplesdir, "h5",recursive = T, full.names = T)

names(files) <- 
  list.dirs(samplesdir,full.names = F)[-c(1)]

bflo_txi.kallisto <- 
  tximport(files, type = "kallisto", txOut = TRUE)
```

## DGE using DESeq2

Load the transcriptome counts in a DESeq2 object:

```{r deseq_object}
bflo_ddsTxi <- 
  DESeqDataSetFromTximport(
    bflo_txi.kallisto, 
    colData = sampleTable,
    design = ~ batch + condition
    )
```

Relevel the factor to first developmental stage.

```{r relevel}
#differential expression analysis
bflo_ddsTxi$condition <- 
  relevel(bflo_ddsTxi$condition, ref = c("01_UFegg"))
```

We estimate the size factors and keep the genes with more than 10 counts throughout the whole dataset. After this we do the actual DGE analysis using Likelihood-Ratio Test as the method because this is time-series data.

```{r DGE}
dds_bflo <- estimateSizeFactors(bflo_ddsTxi)

nc_bflo <- counts(dds_bflo, normalized=TRUE)
filter <- rowSums(nc_bflo >= 10) >= 2
dds_bflo <- dds_bflo[filter,]

dds_bflo <- 
  DESeq(
    dds_bflo,
    parallel = T,
    BPPARAM = BiocParallel::MulticoreParam(12),
    test = "LRT",
    reduced = ~batch
    )
```

We extract the results

```{r}
res_bflo <- results(dds_bflo)
res_bflo
```

And we keep the significant results (top 20000 genes, ranked p-adj.)

```{r}
res_bflo <- res_bflo[order(res_bflo$p.adj),]

cutoff = 0.05 #0.005 # try 0.05?
signif_bflo <- subset(res_bflo, padj<cutoff)
signif_bflo <- signif_bflo[order(signif_bflo$padj),]
signif_bflo <- signif_bflo[1:20000,]
```

We extract the variance-transformed data.

```{r datat ransform}
# raw counts
bflo_counts <- as.data.frame(counts(dds_bflo))
colnames(bflo_counts) <- 
  paste(sampleTable$condition,sampleTable$batch,sep="_")
bflo_counts <- bflo_counts[,order(colnames(bflo_counts))]

# vs-data
bflo_vsd <- as.data.frame(assay(vst(dds_bflo)))
colnames(bflo_vsd) <- 
  paste(sampleTable$condition,sampleTable$batch,sep="_")
bflo_vsd <- bflo_vsd[,order(colnames(bflo_vsd))]

#vs-data of stage-specific genes
rna_bflo <- bflo_vsd[rownames(bflo_vsd) %in% rownames(signif_bflo),]
```

We calculate rows by replicates:

```{r rowmeans_repl}
rna_bflo_q <- qnorm(rna_bflo)

RNA_bflo <- 
  data.frame(
    base::sapply(
      levels(bflo_condition_x),
      rowMeans_by_repl,rna_bflo_q
      )
  )

colnames(RNA_bflo) <- sub("^X","",colnames(RNA_bflo))

```

## Correlation Plot

```{r cor}
bflo_cor <- 
  Heatmap(
    cor(qnorm(bflo_vsd), method = "spearman"),
    name="Amphioxus\n(Spearman)",
    cluster_rows = F,
    cluster_columns = F,
    col = colorRampPalette(rev(brewer.pal(7,"RdYlBu")))(100)
  )

draw(bflo_cor)
```

## PCA plots

```{r PCA}
par(mfrow = c(1,2))
plotPCA(as.matrix(rna_bflo_q), col=rainbow(20), cex=0.75,main="Genes diff. regulated\nduring development\n(separate replicates)") #fix colors
plotPCA(as.matrix(RNA_bflo),col=viridis::viridis(12),cex=0.75,main="Genes diff. regulated\nduring development\n(merged replicates)")
par(mfrow = c(1,1))
```

## Clustering using hierarchical clustering

We will group the stage-specific expressed genes in groups of stage-specific coexpression using hierarchical clustering.

```{r heatmap}
set.seed(1234)
bflo_hc <- hclust(dist(t(scale(t(RNA_bflo)))), method = "ward.D2")
plot(bflo_hc,cex=0.3)
abline(50,0,col="red")

RNA_bflo$clusterID <- cutree(bflo_hc, h=40)
unique(RNA_bflo$clusterID)
```

```{r, fig.width = 4, fig.height = 7.5}
bflo_cl_avgs <-
  aggregate(
    t(scale(t(RNA_bflo[, 1:12]))),
    by = list(cluster = RNA_bflo$clusterID),
    FUN = mean
  )[, -1]

bflo_heat_avg <- 
  Heatmap(
    bflo_cl_avgs,
    name = "gene exp",
    show_row_names = T,
    cluster_rows=T,
    cluster_columns=F,
    column_names_side="top",
    col = diverging_hcl(20,"Red-Green"),
    row_labels = c(1:23)
  )

draw(bflo_heat_avg)
```

Based on visual inspection, we assign new cluster names to give them new ordering:

```{r newcl}
bflo_newcl <- 
  read.table(
    file = "outputs/comparative/reanalysis/bflo/bflo_newcl.tsv",
    header = FALSE,
    col.names = c("newcl","clsuterID")
  )[,c(2,1)]

RNA_bflo$newcl <-
  translate_ids(
    x = RNA_bflo$clusterID,
    dict = bflo_newcl
  )

#We do not need the original hierachical cluster ids anymore
# RNA_bflo$clusterID <- NULL

```

And a final Heatmap for visualisation:

```{r, fig.width = 4, fig.height = 7.5}
set.seed(1234)
RNA_bflo_viz <- RNA_bflo[sample(nrow(RNA_bflo),2000),]
RNA_bflo_viz <- RNA_bflo_viz[order(RNA_bflo_viz$newcl),]
bflo_heat <- 
  Heatmap(
    t(scale(t(RNA_bflo_viz[,1:12]))),
    name = "gene exp",
    # left_annotation= bl_col_annot,
    row_split=RNA_bflo_viz$newcl,
    show_row_names = F,
    cluster_rows=F,
    cluster_columns=F,
    row_title=NULL,
    # left_annotation = 
    #   rowAnnotation(
    #     cluster = anno_block(
    #       gp = gpar(col=NA,fill=NA),
    #       labels = unique(sort(RNA_bflo_viz$newcl)), 
    #       labels_gp = gpar(col = "black", cex=0.7),
    #       labels_just = "right",
    #       labels_offset = unit(1, "npc"),
    #       labels_rot=0
    #       )
    #     ),
  column_names_side="top",
  col = diverging_hcl(20,"Red-Green")
)

draw(bflo_heat)
```

And the average expression profile of the different clusters of stage-specific genes:

```{r avg_profiles}
bflo_cl_avgs <-
  aggregate(
    t(scale(t(RNA_bflo[, 1:12]))),
    by = list(cluster = RNA_bflo$newcl),
    FUN = mean
  )[, -1]

bflo_cl_colors <- 
  colorRampPalette(colors=brewer.pal(11,"Spectral"))(25)

plot(
  0,
  xlim = c(1, 12),
  ylim = c(-2, 3),
  type = "n",
  main = "Average Expression Profile of\nStage-Specific Gene Clusters",
  xlab = "stage of development",
  ylab = "z-score Expression"
)
for (i in 1:nrow(bflo_cl_avgs)) {
  lines(c(1:12),
        bflo_cl_avgs[i,],
        type = "l",
        col = bflo_cl_colors[i],
        lwd = 2
  )
}
legend(
  x = 1,
  y = 3,
  legend = rownames(bflo_cl_avgs),
  fill = bflo_cl_colors,
  ncol = 5,
  cex = 0.6
)
```

```{r bflo_avg_heat, fig.width = 3, fig.height = 5}
bflo_heat_avg <- 
  Heatmap(
    bflo_cl_avgs,
    name = "gene exp",
    show_row_names = T,
    row_labels = seq(nrow(bflo_cl_avgs)),
    cluster_rows=F,
    cluster_columns=F,
    column_names_side="top",
    col = diverging_hcl(20,"Red-Green")
  )

draw(bflo_heat_avg)
```

```{r}
pdf("graphics/bflo_clusters.pdf", height = 5, width = 3)
draw(bflo_heat_avg)
dev.off()
```


## Save the data

We finally translate the gene ids to the same ones used in OMA:

```{r translate_ids}
rownames(bflo_counts) <- translate_ids(x = rownames(bflo_counts), dict = bf_bflo, return.missing = TRUE)
rownames(bflo_vsd) <- translate_ids(x = rownames(bflo_vsd), dict = bf_bflo, return.missing = TRUE)
rownames(rna_bflo) <- translate_ids(x = rownames(rna_bflo), dict = bf_bflo, return.missing = TRUE)
rownames(rna_bflo_q) <- translate_ids(x = rownames(rna_bflo_q), dict = bf_bflo, return.missing = TRUE)
rownames(RNA_bflo) <- translate_ids(x = rownames(RNA_bflo), dict = bf_bflo, return.missing = TRUE)
```

And we create a table of gene <-> cluster for later use in our comparative analyses:

```{r bflo_cl}
bflo_cl <- 
  data.frame(
    id = rownames(RNA_bflo),
    cID = RNA_bflo$newcl
  )
```

And we save the data for TF analysis and for pairwise comparisons:

```{r save}
save(
  bflo_counts,
  bflo_vsd,
  rna_bflo,
  rna_bflo_q,
  RNA_bflo,
  bflo_cl,
  bflo_cl_avgs,
  bf_bflo,
  file = "outputs/rda/bflo_reanalysis.rda"
)
```