---
title: "Ptychodera Cisreg Development: pairwise species comparisons"
author: "Alberto Perez-Posada"
date: "4/22/2023"
output: html_document
---

```{r setup, include=FALSE}
dir <- '/home/ska/aperpos/projects/ptychodera_cisreg_development/'
fcha <- function(){ gsub("-","",Sys.Date()) }

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = dir)
options(scipen=999)
Sys.setenv(VROOM_CONNECTION_SIZE=5000072)
```

## About

In this markdown, we will do a comparison of the transcriptomes during development of Ptychodera, Amphixous, and Purple Sea Urchin.

For this, we will use a solution of custom R functions and wrappers that we have named "comparABle" (kudos to @agilgal for the name!).

## Load libraries

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(stringr)
library(BiocGenerics)
library(EDASeq)
library(DESeq2)
library(tximport)
library(limma)
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
require(data.table)
require(stringr)
require(tidytree)
require(ggtree)
require(rvcheck)
require(treeio)
require(dendextend)
require(phangorn)
require(phytools)
require(ComplexHeatmap)
require(foreach)
require(doParallel)
require(colorspace)
library(topGO)
```

## Load functions

```{r load_functions, message = FALSE}
source("code/r_code/r_functions/sourcefolder.R")

sourceFolder(
  "code/r_code/r_functions",
  recursive = TRUE
  )

sourceFolder(
  "code/r_code/r_general",
  recursive = TRUE
  )
```

## Load Data

We will first load the data of our three species:

```{r dataload_spp}
# Ptychodera analyses
load("outputs/rda/deseq2.rda")
# Amphioxus reanalysis
load("outputs/rda/blan_reanalysis.rda")
# Purple Sea Urchin reanalysis
load("outputs/rda/spur_reanalysis.rda")
```

We will follow by loading all the gene family data that we need. This is the gene/gfam lookup table (extracted from the OrthoXML output file from OMA) and the respective table to transform the gene indexes into the actual gene id of every species. 

```{r dataload_gfams}
gene_gfam <- read.delim2("outputs/comparative/20240404_oma/oma_omaid_gfam.tsv", header = TRUE)
colnames(gene_gfam) <- c("id","gfam")
gene_gfam$id <- gsub("\\.p[0-9]+","",gene_gfam$id)
```

This `gene_gfam` file will be used later on as proxy to compare developmental stages.

```{r gene_gfam}
head(gene_gfam)
```

```{r dataload_1to1_oma_orthologues}
oma_pfla_blan <- read.delim2("outputs/comparative/20240404_oma/one_to_one_orthologues/1to1_pfla_blan.tsv", header = FALSE)
oma_pfla_blan$V2 <- gsub(" ","",oma_pfla_blan$V2)
oma_pfla_spur <- read.delim2("outputs/comparative/20240404_oma/one_to_one_orthologues/1to1_pfla_spur.tsv", header = FALSE)
oma_pfla_spur$V2 <- gsub(" ","",oma_pfla_spur$V2)
```

```{r geneage prep}
load("outputs/rda/geneage.rda")
```

For more functional annotation, we will use the COG functional categories and the GO terms of Ptychodera.

```{r gos and cogs prep}
pfla_cogs <- read.table(
  "outputs/functional_annotation/COGs/pfla_cogs.tsv",
  col.names = c("id","cog")
)

# GO terms
pfla_id2go <- 
  readMappings(
    "outputs/functional_annotation/go_blast2go/GO_annotation.txt"
  )
```

## Ptychodera and Amphioxus

We will first compare the transcriptomes and stage-specific clusters of Ptychodera and Amphioxus. For this, we will define a number of objects that will enter into our custom wrapper function `comparABle`. The most important of them being:

 - Objects **a** and **b** are the expression profile datasets (gene x condition/stage/etc) of species A and B.
 - Object **o** is a 1-to-1 orthologue association file between genes of species A and B (e.g. BLAST reciprocal best hits, OMA 1:1 orthologues, ...)
 - Object **f** is a "gene family"/orthogroup file associating genes from species A and B to "gene families"
 - Objects **ma**, **mb** are a gene classification system. For each species a and b, a file associating genes to gene moules (e.g. k-means clusters, hierarchical clusters, WGCNA modules, etc.)
 - Object **g** is a gene age file for the gene families, and the genes, of each species.

In addition, we will also pass it a list of gene ontologies, a COG functional category association file, and a list of gene ages that are shared between species A and B.

```{r pfla_blan dataprep}
# Expression data
a = pfla_rna_counts
b = blan_counts

# Samples for rowmeans_by repl
a_samples = levels(condition_x)
b_samples = unique(sub("_.$", "", colnames(b)))

# Family/orthology data
o = unique(oma_pfla_blan)
f = gene_gfam[grep("TCONS|BRALA",gene_gfam$id),]

# Module/Cluster information
ma = data.frame(
  id = rownames(pfla_rna_dev),
  module = pfla_rna_dev$cID
)

mb = blan_cl
colnames(mb) <- c("id","module")

# Gene Age
ga = pfla_age[,c(1,3)]
colnames(ga) = c("id","age")

# COG
cog_a <- pfla_cogs

# GOs
a_universe = rownames(vsd_allgen)
a_id2go = pfla_id2go

# Common Evo Nodes
common_evo_nodes = unique(ga$age)[!(unique(ga$age) %in% c("6_ambulacraria","7_hemichordata","8_Pfla"))]
```

Below we will run the comparABle wrapper. Briefly, what this wrapper does is:

 - Normalise the expression datasets
 - Associate the datasets using the association file of 1-to-1 orthologues
 - Perform a principal component analysis with the merge of the two datasets
 - Perform pairwise correlations and distance metrics between conditions (in our case, developmental stages) of datasets to see their relationships of similarity. By default: Pearson Correlation, Spearman Correlation, and Jensen-Shannon Divergence.
 - Identify the genes showcasing highes values of correlations across a number of pre-determined conditions (in our case, stages), and perform Gene Age and Gene Ontology enrichment analysis
 - Perform a co-occurrence analysis in the merge of the two datasets (pairwise correlations, hierarchical clustering, and bootstrapping with downsampling)
 - Perform Orthology Overlap analysis (hypergeometric and binomial) in pairwise correlations between gene sets (==stage-specific clusters) of each species.
 - Identify the genes of the gene families that appear significantly enriched across pairs of gene sets, and perform Gene Age, Gene Ontology, and Functional Category enrichment analyses.
 
 This can take some time.

```{r pfla_blan comparABle, fig.width = 7, fig.height = 6, warning = FALSE, message = FALSE}
PFLA_BLAN_COMPARISON <- comparABle(
  a_name = "P.flava",
  b_name = "B.lanceolatum",
  a = a,
  b = b,
  o = o,
  f = f,
  ma = ma,
  mb = mb,
  ga = ga,
  gb = gb,
  cog_a = cog_a,
  cog_b = cog_b,
  a_samples = a_samples,
  b_samples = b_samples,
  highlyvariable = FALSE,
  cooc_p = 0.05,
  cooc_h = c(0.70,0.95),
  cooc_cor_method = "pearson",
  a_universe = a_universe,
  a_id2go = a_id2go,
  common_evo_nodes = common_evo_nodes,
  sep = ",\ "
)
```

After running this we can start exploring the comparisons. First we will tidy up the pairwise correlation data and add names to the rows and columns of the jensen-shannon distance matrix.

Here are the heatmaps of pairwise correlations between Ptychodera and Amphioxus.

```{r corr_heatmaps, fig.width = 10, fig.height = 3.5}
plot_cors(PFLA_BLAN_COMPARISON$pairwise_correlations)
```
A quick sanity check that this is consistently observed regardles of genes by doing some bootstraping and checkin the mean observed JSD:
```{r js_mean_pfla_blan, fig.width = 3.5, fig.height = 3.5}
js_mean_pfla_blan <- jsd_with_subsampling(
  a_o = PFLA_BLAN_COMPARISON$merged_data$a_o , 
  b_o = PFLA_BLAN_COMPARISON$merged_data$b_o, 
  n = 1000, p = 0.25
)

#relativise to min and max values
js_mean_rel <- relativise(js_mean_pfla_blan$mean)

h3_avg_rel <- Heatmap(js_mean_rel,cluster_rows = F, cluster_columns = F, show_row_names = TRUE, name = "JSD", col = brewer.pal(10,"RdBu"))
draw(h3_avg_rel)
```

Here is the average expression profile of genes more expressed in the most similar stages between Ptychodera an Amphioxus, including GO terms and age enrichment:

```{r pfla_spur plots, fig.width = 10, fig.height = 20, message = FALSE, warning = FALSE}
# Common genes, highly correlated
p1 <- plot_hicor_genes(
  scatter = PFLA_BLAN_COMPARISON$high_corr_genes$pairwise_data$hicor_topgenes$cor_05_MG__08_18h$plot_topgenes,
  GO_plot = PFLA_BLAN_COMPARISON$high_corr_genes$GOs$GOplot$cor_05_MG__08_18h,
  age_table = PFLA_BLAN_COMPARISON$high_corr_genes$age$cor_05_MG__08_18h$AgeperModule,
  age_enrichemnt_hm = PFLA_BLAN_COMPARISON$high_corr_genes$age$cor_05_MG__08_18h$heatmap
)
p2 <- plot_hicor_genes(
  scatter = PFLA_BLAN_COMPARISON$high_corr_genes$pairwise_data$hicor_topgenes$cor_04_EG__08_18h$plot_topgenes,
  GO_plot = PFLA_BLAN_COMPARISON$high_corr_genes$GOs$GOplot$cor_04_EG__08_18h,
  age_table = PFLA_BLAN_COMPARISON$high_corr_genes$age$cor_04_EG__08_18h$AgeperModule,
  age_enrichemnt_hm = PFLA_BLAN_COMPARISON$high_corr_genes$age$cor_04_EG__08_18h$heatmap
)
p3 <- plot_hicor_genes(
  scatter = PFLA_BLAN_COMPARISON$high_corr_genes$pairwise_data$hicor_topgenes$cor_10_Me__14_Premet$plot_topgenes,
  GO_plot = PFLA_BLAN_COMPARISON$high_corr_genes$GOs$GOplot$cor_10_Me__14_Premet,
  age_table = PFLA_BLAN_COMPARISON$high_corr_genes$age$cor_10_Me__14_Premet$AgeperModule,
  age_enrichemnt_hm = PFLA_BLAN_COMPARISON$high_corr_genes$age$cor_10_Me__14_Premet$heatmap
)

plot_grid(p1,p2,p3,ncol = 1)
```

Here is the co-occurrence analysis showcasing the Amphioxus stages most similar to Ptychodera gastrulation are the stages corresponding to neurulation. 

```{r pfla_blan_cooc, fig.width = 7, fig.height = 6}
Heatmap(
  name = "co-occurrence",
  PFLA_BLAN_COMPARISON$coocurrence_analysis$cooccurrence,
  cluster_rows = PFLA_BLAN_COMPARISON$coocurrence_analysis$tree,
  cluster_columns = PFLA_BLAN_COMPARISON$coocurrence_analysis$tree,
  col = sequential_hcl(10,"YlOrRd", rev = TRUE)
)
```
```{r pfla_blan_cooc_noclust, fig.width = 4.5, fig.height = 4}
Heatmap(
  name = "co-occurrence",
  PFLA_BLAN_COMPARISON$coocurrence_analysis$cooccurrence[1:16,17:28],
  cluster_rows = F,
  cluster_columns = F,
  col = sequential_hcl(10,"YlOrRd", rev = TRUE)
)
```


Here the orthology overlap strategy showcasing similarities at the gene family usage between hemichordate larval stages and amphioxus post-gastrulation development

```{r orthology_overlap_strategy, fig.width = 11.5, fig.height = 4}
pf_avg_hm+
PFLA_BLAN_COMPARISON$plots$orthology_overlap_binomial_hm+
PFLA_BLAN_COMPARISON$plots$orthology_overlap_hypgeom_hm
```

Here is a quick overview of the stats of the hypergeometric and binomial tests for pairwise comparisons:

```{r}
summary(PFLA_BLAN_COMPARISON$orthology_overlap_modules$pairwise_module_comparison$stats)
```

These is a quick overview at the number of genes in enriched gene families between pairs of comparisons:

```{r top fams}
# Group by module and summarize
commongenes <- PFLA_BLAN_COMPARISON$orthology_overlap_modules$
  genes_in_common_fams$commonfams$table_a_common %>%
  dplyr::group_by(module) %>%
  dplyr::summarize(
    numgenes = n(),
    genes = paste(ifelse(n() > 3, paste0(paste(head(id, 3), collapse = ",")," ..."), paste(id, collapse = ", ")), collapse = ", ")
  )

# Rename columns
colnames(commongenes) <- c("module", "numgenes", "genes")

commongenes
```

And here the combination of gene age and functional categories enrichment:

```{r keygenes commonfams age, fig.width = 9, fig.height = 5}
PFLA_BLAN_COMPARISON$orthology_overlap_modules$
  genes_in_common_fams$commonfams$
    age_a_common$heatmap +
PFLA_BLAN_COMPARISON$orthology_overlap_modules$
  genes_in_common_fams$commonfams$
    cog_a_comon$heatmap
```

And here the gene ontology enrichment of enriched gene families between pairs of modules.

```{r}
# Add a new column with the name of the original data frame
keygenes_commonfams_GO <- 
  bind_rows(
    PFLA_BLAN_COMPARISON$orthology_overlap_modules$
      genes_in_common_fams$commonfams$go_a_common$GOtable, 
    .id = "pair_gfams"
    )

# Rename the new column
colnames(keygenes_commonfams_GO)[1] <- "pair_gfams"

keygenes_commonfams_GO$logp <- 
  -log10(as.numeric(keygenes_commonfams_GO$classicFisher))

keygenes_commonfams_GO$logp[is.na(keygenes_commonfams_GO$logp)] <- 
  max(keygenes_commonfams_GO$logp, na.rm = T)

keygenes_commonfams_GO$obsexp <- 
  as.numeric(keygenes_commonfams_GO$Significant/keygenes_commonfams_GO$Expected)

keygenes_commonfams_GO <- 
  keygenes_commonfams_GO[
    order(
      keygenes_commonfams_GO$pair_gfams,
      keygenes_commonfams_GO$Term
      ),
    ]

keygenes_commonfams_GO$Term <- 
  factor(
    keygenes_commonfams_GO$Term, 
    levels = unique(keygenes_commonfams_GO$Term)
    )
```

The GO plot here:

```{r pfla_blan_keygenescommonfams GOterms_plot, fig.width = 10, fig.height = 50}
gg <- keygenes_commonfams_GO  %>% 
  ggplot(
    aes(x=pair_gfams, y = Term, size = Significant, color = logp)
    ) + geom_point() +
  theme_bw() + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8 )
    )+ 
  theme(
    axis.text.y.left = element_text(hjust = 1, size = 8 ),
    legend.title=element_text(size=8 ),
    legend.text=element_text(size=8 )
    )+ 
  scale_size_continuous(
    name = "Genes affected",
    range = c(2,6)
    ) + scale_color_viridis("-log10(P-value)") +
  xlab("")+ylab("")

gg
```

## Ptychodera and Purple Sea Urchin

We will proceed the same as before using sea urchin in this case. Again, we define all the data necessary for the comparABle wrapper:

```{r pfla_spur dataprep}
# Expression data
a = pfla_rna_counts
b = spur_vsd

# Samples for rowmeans_by repl
a_samples = levels(condition_x)
b_samples = unique(sub("_.$", "", colnames(b)))

# Family/orthology data
o = unique(oma_pfla_spur)
f = gene_gfam[grep("TCONS|STRPU",gene_gfam$id),]

# Module/Cluster information
ma = data.frame(
  id = rownames(pfla_rna_dev),
  module = pfla_rna_dev$cID
)

mb = spur_cl
colnames(mb) <- c("id","module")

# Gene Age
ga = pfla_age[,c(1,3)]
colnames(ga) = c("id","age")

# COG
cog_a <- pfla_cogs

# GOs
a_universe = rownames(vsd_allgen)
a_id2go = pfla_id2go

# Common Evo Nodes
common_evo_nodes = unique(ga$age)[!(unique(ga$age) %in% c("7_hemichordata","8_Pfla"))]
```

And we run comparABle:

```{r pfla_spur comparABle, fig.width = 7, fig.height = 6, warning = FALSE, message = FALSE}
PFLA_SPUR_COMPARISON <- comparABle(
  a_name = "P.flava",
  b_name = "S.purpuratus",
  a = a,
  b = b,
  o = o,
  f = f,
  ma = ma,
  mb = mb,
  ga = ga,
  gb = gb,
  cog_a = cog_a,
  cog_b = cog_b,
  a_samples = a_samples,
  b_samples = b_samples,
  highlyvariable = FALSE,
  cooc_p = 0.25,
  cooc_h = c(0.6,0.9),
  cooc_n = 10000,
  cooc_cor_method = "pearson",
  a_universe = a_universe,
  a_id2go = a_id2go,
  common_evo_nodes = common_evo_nodes,
  sep = ",\ "
)
```

The heatmaps of pairwise correlations and jensen-shannon divergence between Ptychodera and Sea Urchin:

```{r , fig.width = 10, fig.height = 3.5}
plot_cors(PFLA_SPUR_COMPARISON$pairwise_correlations)
```

A quick sanity check that this is consistently observed regardles of genes by doing some bootstraping and checkin the mean observed JSD:
```{r js_mean_pfla_spur, fig.width = 3.5, fig.height = 3.5}
js_mean_pfla_spur <- jsd_with_subsampling(
  a_o = PFLA_SPUR_COMPARISON$merged_data$a_o , 
  b_o = PFLA_SPUR_COMPARISON$merged_data$b_o, 
  n = 1000, p = 0.25
)

#relativise to min and max values
js_mean_rel <- relativise(js_mean_pfla_spur$mean)

h3_avg_rel <- Heatmap(js_mean_rel,cluster_rows = F, cluster_columns = F, show_row_names = TRUE, name = "JSD", col = brewer.pal(10,"RdBu"))
draw(h3_avg_rel)
```

Here is the average expression profile of genes more expressed in the most similar stages between Ptychodera an Sea Urchin, including GO terms and age enrichment:

```{r pfla_blan plots, fig.width = 10, fig.height = 20, message = FALSE, warning = FALSE}
# Common genes, highly correlated
p1 <- plot_hicor_genes(
  scatter = PFLA_SPUR_COMPARISON$high_corr_genes$pairwise_data$hicor_topgenes[[1]]$plot_topgenes,
  GO_plot = PFLA_SPUR_COMPARISON$high_corr_genes$GOs$GOplot[[1]],
  age_table = PFLA_SPUR_COMPARISON$high_corr_genes$age[[1]]$AgeperModule,
  age_enrichemnt_hm = PFLA_SPUR_COMPARISON$high_corr_genes$age[[1]]$heatmap
)
p2 <- plot_hicor_genes(
  scatter = PFLA_SPUR_COMPARISON$high_corr_genes$pairwise_data$hicor_topgenes[[2]]$plot_topgenes,
  GO_plot = PFLA_SPUR_COMPARISON$high_corr_genes$GOs$GOplot[[2]],
  age_table = PFLA_SPUR_COMPARISON$high_corr_genes$age[[2]]$AgeperModule,
  age_enrichemnt_hm = PFLA_SPUR_COMPARISON$high_corr_genes$age[[2]]$heatmap
)
p3 <- plot_hicor_genes(
  scatter = PFLA_SPUR_COMPARISON$high_corr_genes$pairwise_data$hicor_topgenes[[3]]$plot_topgenes,
  GO_plot = PFLA_SPUR_COMPARISON$high_corr_genes$GOs$GOplot[[3]],
  age_table = PFLA_SPUR_COMPARISON$high_corr_genes$age[[3]]$AgeperModule,
  age_enrichemnt_hm = PFLA_SPUR_COMPARISON$high_corr_genes$age[[3]]$heatmap
)

plot_grid(p1,p2,p3,ncol = 1)
```

And the co-ocurrence matrix shocasing ptychodera gastrulation is similar to sea urchin gastrulation as well as similarities between larval stages

```{r pfla_spur_cooc, fig.width = 7, fig.height = 6}
Heatmap(
  name = "co-occurrence",
  PFLA_SPUR_COMPARISON$coocurrence_analysis$cooccurrence,
  cluster_rows = PFLA_SPUR_COMPARISON$coocurrence_analysis$tree,
  cluster_columns = PFLA_SPUR_COMPARISON$coocurrence_analysis$tree,
  col = sequential_hcl(10,"YlOrRd", rev = TRUE)
)
```

```{r pfla_spur_cooc_noclust, fig.width = 4.5, fig.height = 4}
Heatmap(
  name = "co-occurrence",
  PFLA_SPUR_COMPARISON$coocurrence_analysis$cooccurrence[1:16,17:33],
  cluster_rows = F,
  cluster_columns = F,
  col = sequential_hcl(10,"YlOrRd", rev = TRUE)
)
```

Below the orthology overlap strategy showcasing similarities at the gene family usage between species:

```{r , fig.width = 11.5, fig.height = 4}
pf_avg_hm+
PFLA_SPUR_COMPARISON$plots$orthology_overlap_binomial_hm+
PFLA_SPUR_COMPARISON$plots$orthology_overlap_hypgeom_hm
```

An overview of the stats of these tests:

```{r}
str(PFLA_SPUR_COMPARISON$orthology_overlap_modules$pairwise_module_comparison$stats)
```

A look at the number of genes belonging to enriched gene families between comparisons

```{r}
# Group by module and summarize
commongenes <- PFLA_SPUR_COMPARISON$orthology_overlap_modules$
  genes_in_common_fams$commonfams$table_a_common %>%
  dplyr::group_by(module) %>%
  dplyr::summarize(
    numgenes = n(),
    genes = paste(ifelse(n() > 3, paste0(paste(head(id, 3), collapse = ",")," ..."), paste(id, collapse = ", ")), collapse = ", ")
  )

# Rename columns
colnames(commongenes) <- c("module", "numgenes", "genes")

commongenes
```

The gene age enrichment and functional category enrichment of these genes:

```{r , fig.width = 9, fig.height = 5}
PFLA_SPUR_COMPARISON$orthology_overlap_modules$
  genes_in_common_fams$commonfams$
    age_a_common$heatmap +
PFLA_SPUR_COMPARISON$orthology_overlap_modules$
  genes_in_common_fams$commonfams$
    cog_a_comon$heatmap
```

And the GO terms of these genes:

```{r}
# Add a new column with the name of the original data frame
keygenes_commonfams_GO <- 
  bind_rows(
    PFLA_SPUR_COMPARISON$orthology_overlap_modules$
      genes_in_common_fams$commonfams$go_a_common$GOtable, 
    .id = "pair_gfams"
    )

# Rename the new column
colnames(keygenes_commonfams_GO)[1] <- "pair_gfams"

keygenes_commonfams_GO$logp <- 
  -log10(as.numeric(keygenes_commonfams_GO$classicFisher))

keygenes_commonfams_GO$logp[is.na(keygenes_commonfams_GO$logp)] <- 
  max(keygenes_commonfams_GO$logp, na.rm = T)

keygenes_commonfams_GO$obsexp <- 
  as.numeric(keygenes_commonfams_GO$Significant/keygenes_commonfams_GO$Expected)

keygenes_commonfams_GO <- 
  keygenes_commonfams_GO[
    order(
      keygenes_commonfams_GO$pair_gfams,
      keygenes_commonfams_GO$Term
      ),
    ]

keygenes_commonfams_GO$Term <- 
  factor(
    keygenes_commonfams_GO$Term, 
    levels = unique(keygenes_commonfams_GO$Term)
    )
```

The plot of the GO enrichment analysis:

```{r pfla_spur_keygenescommonfams GOterms_plot, fig.width = 10, fig.height = 50}
gg <- keygenes_commonfams_GO  %>% 
  ggplot(
    aes(x=pair_gfams, y = Term, size = Significant, color = logp)
    ) + geom_point() +
  theme_bw() + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8 )
    )+ 
  theme(
    axis.text.y.left = element_text(hjust = 1, size = 8 ),
    legend.title=element_text(size=8 ),
    legend.text=element_text(size=8 )
    )+ 
  scale_size_continuous(
    name = "Genes affected",
    range = c(2,6)
    ) + scale_color_viridis("-log10(P-value)") +
  xlab("")+ylab("")
```

```{r, fig.height = 14, fig.width = 9}
stages_pf_sp = 
  list(
    a = c("04_EG","04_EG"),
    b = c("03_MeBl_24h1","03_MeBl_24h2")
  )

pfla_spur_hicor_genes <- get_high_cor_genes(
  mat = PFLA_SPUR_COMPARISON$pairwise_correlations$js,
  a_o = PFLA_SPUR_COMPARISON$merged_data$a_o,
  b_o = PFLA_SPUR_COMPARISON$merged_data$b_o,
  stages = stages_pf_sp,
  o = PFLA_SPUR_COMPARISON$input$o
)

pfla_spur_hicor_genes_GOs <- 
  getGOs(
    genelist = 
      lapply(
        pfla_spur_hicor_genes$hicor_topgenes, 
        function(sub_list) {
          setNames(sub_list$top_genes$a, names(sub_list))
        }),
    gene_universe = a_universe,
    gene2GO = a_id2go,
    max_terms = 10
  )

print("Common genes in Correlations (age)")
pfla_spur_hicor_genes_age <- 
  lapply(
    lapply(
      pfla_spur_hicor_genes$hicor_topgenes,
      function(sub_list) {
        data.frame(
          id = a_universe,
          module = 
            ifelse(
              a_universe %in% sub_list$top_genes$a, "common","not_common"
            )
        )
      }
    ),
    function(x){
      gene_age_enrichment(
        x_modules = x,
        x_age = ga[ga$age %in% common_evo_nodes,]
      )
    }
  )


# Common genes, highly correlated
p1 <- plot_hicor_genes(
  scatter = pfla_spur_hicor_genes$hicor_topgenes[[1]]$plot_topgenes,
  GO_plot = pfla_spur_hicor_genes_GOs$GOplot[[1]],
  age_table = pfla_spur_hicor_genes_age[[1]]$AgeperModule,
  age_enrichemnt_hm = pfla_spur_hicor_genes_age[[1]]$heatmap
)
p2 <- plot_hicor_genes(
  scatter = pfla_spur_hicor_genes$hicor_topgenes[[2]]$plot_topgenes,
  GO_plot = pfla_spur_hicor_genes_GOs$GOplot[[2]],
  age_table = pfla_spur_hicor_genes_age[[2]]$AgeperModule,
  age_enrichemnt_hm = pfla_spur_hicor_genes_age[[2]]$heatmap
)

plot_grid(
  p1,
  p2,
  ncol = 1)
```

## Alternative method for 1-to-1 orthologs: Blast best reciprocal hits

# Comparison of 1-to-1 TFs using blast best RBHs

```{r dataload_1to1_rbh}
rbh_pfla_blan <- read.delim2("outputs/comparative/rbh/pfla_blan_rbh.tsv", header = FALSE)
rbh_pfla_blan[,2] <- gsub(" ","",rbh_pfla_blan[,2])
rbh_pfla_blan[,1] <- gsub("\\.p[0-9]+","",rbh_pfla_blan[,1])

rbh_pfla_spur <- read.delim2("outputs/comparative/rbh/pfla_spur_rbh.tsv", header = FALSE)
rbh_pfla_spur[,2] <- gsub(" ","",rbh_pfla_spur[,2])
rbh_pfla_spur[,1] <- gsub("\\.p[0-9]+","",rbh_pfla_spur[,1])
```

## Ptychodera and Amphioxus

```{r pfla_blan plot_cors, fig.width = 10, fig.height = 3.5}
# Expression data, using all genes. No need for transformation
a = PFLA_BLAN_COMPARISON$input$a
b = PFLA_BLAN_COMPARISON$input$b

# Family/orthology data
o = unique(rbh_pfla_blan)

# MERGE
o = pair2id(o)

pf_bl_merge_ab <- mergedata(a,b,o)

# CORRELATIONS
pf_bl_rbh_cors_allgen <- rawcorsp(pf_bl_merge_ab$a_o,pf_bl_merge_ab$b_o) # FIX JSD

plot_cors(pf_bl_rbh_cors_allgen)
```


## Ptychodera and Sea Urchin

```{r pfla_spur plot_cors, fig.width = 10, fig.height = 3.5}
# Expression data
a = PFLA_SPUR_COMPARISON$input$a
b = PFLA_SPUR_COMPARISON$input$b

# Family/orthology data
o = unique(rbh_pfla_spur)
o = pair2id(o)

# MERGE
pf_sp_merge_ab <- mergedata(a,b,o)

# CORRELATIONS
print("Correlations")
pf_sp_rbh_cors_allgen <- rawcorsp(pf_sp_merge_ab$a_o,pf_sp_merge_ab$b_o) # FIX JSD

plot_cors(pf_sp_rbh_cors_allgen)
```

## Saving the data

We will save these two objects for further analysis and reference:

```{r save}
save(
  PFLA_BLAN_COMPARISON,
  PFLA_SPUR_COMPARISON,
  file = "outputs/rda/species_comparison.rda"
)
```

## Figures for the paper

These are just calls to PDFs for plotting the figures we have generated.

```{r echo = FALSE, eval = FALSE}
pf_sp_js_rel <- relativise(js_mean_pfla_spur$mean)

pf_sp_js_hm <- Heatmap(
  pf_sp_js_rel,
  cluster_rows = F,
  cluster_columns = F,
  show_row_names = TRUE,
  name = "JSD",
  col = brewer.pal(10,"RdBu"),
  left_annotation = devstages_ha_rows(),
  top_annotation = quick_ha(colnames(pf_sp_js_rel),"Purple-Orange", rev = TRUE)
  )

pf_bl_js_rel <- relativise(js_mean_pfla_blan$mean)

pf_bl_js_hm <- Heatmap(
  pf_bl_js_rel,
  cluster_rows = F,
  cluster_columns = F,
  show_row_names = TRUE,
  name = "JSD",
  col = brewer.pal(10,"RdBu"),
  left_annotation = devstages_ha_rows(),
  top_annotation = quick_ha(colnames(pf_bl_js_rel),"Purple-Yellow", rev = TRUE)
)

pf_sp_cooc <- 
  grid.grabExpr(draw(Heatmap(
    name = "co-occurrence",
    PFLA_SPUR_COMPARISON$coocurrence_analysis$cooccurrence[1:16,17:33],
    cluster_rows = F,
    cluster_columns = F,
    col = sequential_hcl(10,"YlOrRd", rev = TRUE),
    left_annotation = devstages_ha_rows(),
    top_annotation = quick_ha(colnames(PFLA_SPUR_COMPARISON$coocurrence_analysis$cooccurrence[1:16,17:33]),"Purple-Orange", rev = TRUE)
  )))

pf_bl_cooc <- 
  grid.grabExpr(draw(Heatmap(
    name = "co-occurrence",
    PFLA_BLAN_COMPARISON$coocurrence_analysis$cooccurrence[1:16,17:28],
    cluster_rows = F,
    cluster_columns = F,
    col = sequential_hcl(10,"YlOrRd", rev = TRUE),
    left_annotation = devstages_ha_rows(),
    top_annotation = quick_ha(colnames(PFLA_BLAN_COMPARISON$coocurrence_analysis$cooccurrence[1:16,17:28]),"Purple-Yellow", rev = TRUE)
  )))

pf_sp_ol <-
  grid.grabExpr(draw(PFLA_SPUR_COMPARISON$plots$orthology_overlap_hypgeom_hm))

pf_bl_ol <-
  grid.grabExpr(draw(PFLA_BLAN_COMPARISON$plots$orthology_overlap_hypgeom_hm))

pf_sp_funcat <-
  grid.grabExpr(draw(PFLA_SPUR_COMPARISON$orthology_overlap_modules$
    genes_in_common_fams$commonfams$
    cog_a_comon$heatmap))

pf_bl_funcat <-
  grid.grabExpr(draw(PFLA_BLAN_COMPARISON$orthology_overlap_modules$
    genes_in_common_fams$commonfams$
    cog_a_comon$heatmap))


hm1 <- grid.grabExpr(draw(pf_sp_js_hm))
hm2 <- grid.grabExpr(draw(pf_bl_js_hm))

pdf(
  file = "graphics/4A_B.pdf",
  height = 4.5,
  width = 7
)
draw(pf_sp_js_hm+pf_bl_js_hm)
dev.off()

save_plot(
  file = "graphics/4C_D.pdf",
  base_height = 5,
  base_width = 3,
  plot_grid(
      pfla_spur_hicor_genes$hicor_topgenes[[2]]$plot_topgenes,
      PFLA_BLAN_COMPARISON$high_corr_genes$pairwise_data$hicor_topgenes[[2]]$plot_topgenes,
      ncol = 1)
)
pdf(
  file = "graphics/4E.pdf",
  height = 8,
  width = 5
)
draw(
  PFLA_SPUR_COMPARISON$orthology_overlap_modules$
    genes_in_common_fams$commonfams$
    cog_a_comon$heatmap %v% 
  PFLA_BLAN_COMPARISON$orthology_overlap_modules$
    genes_in_common_fams$commonfams$
    cog_a_comon$heatmap
  )
dev.off()

# Common genes, highly correlated PF_SPUR
p1 <- plot_hicor_genes(
  scatter = pfla_spur_hicor_genes$hicor_topgenes[[1]]$plot_topgenes,
  GO_plot = pfla_spur_hicor_genes_GOs$GOplot[[1]],
  age_table = pfla_spur_hicor_genes_age[[1]]$AgeperModule,
  age_enrichemnt_hm = pfla_spur_hicor_genes_age[[1]]$heatmap
)
p2 <- plot_hicor_genes(
  scatter = pfla_spur_hicor_genes$hicor_topgenes[[2]]$plot_topgenes,
  GO_plot = pfla_spur_hicor_genes_GOs$GOplot[[2]],
  age_table = pfla_spur_hicor_genes_age[[2]]$AgeperModule,
  age_enrichemnt_hm = pfla_spur_hicor_genes_age[[2]]$heatmap
)

save_plot(
  file = "graphics/pfla_spur_hicor_eg_mebl.pdf",
  base_height = 10,
  base_width = 10,
  plot_grid(
    p1,
    p2,
    ncol = 1
    )
  )

# Common genes, highly correlated PF_BL
p1 <- plot_hicor_genes(
  scatter = PFLA_BLAN_COMPARISON$high_corr_genes$pairwise_data$hicor_topgenes$cor_05_MG__08_18h$plot_topgenes,
  GO_plot = PFLA_BLAN_COMPARISON$high_corr_genes$GOs$GOplot$cor_05_MG__08_18h,
  age_table = PFLA_BLAN_COMPARISON$high_corr_genes$age$cor_05_MG__08_18h$AgeperModule,
  age_enrichemnt_hm = PFLA_BLAN_COMPARISON$high_corr_genes$age$cor_05_MG__08_18h$heatmap
)
p2 <- plot_hicor_genes(
  scatter = PFLA_BLAN_COMPARISON$high_corr_genes$pairwise_data$hicor_topgenes$cor_04_EG__08_18h$plot_topgenes,
  GO_plot = PFLA_BLAN_COMPARISON$high_corr_genes$GOs$GOplot$cor_04_EG__08_18h,
  age_table = PFLA_BLAN_COMPARISON$high_corr_genes$age$cor_04_EG__08_18h$AgeperModule,
  age_enrichemnt_hm = PFLA_BLAN_COMPARISON$high_corr_genes$age$cor_04_EG__08_18h$heatmap
)
p3 <- plot_hicor_genes(
  scatter = PFLA_BLAN_COMPARISON$high_corr_genes$pairwise_data$hicor_topgenes$cor_10_Me__14_Premet$plot_topgenes,
  GO_plot = PFLA_BLAN_COMPARISON$high_corr_genes$GOs$GOplot$cor_10_Me__14_Premet,
  age_table = PFLA_BLAN_COMPARISON$high_corr_genes$age$cor_10_Me__14_Premet$AgeperModule,
  age_enrichemnt_hm = PFLA_BLAN_COMPARISON$high_corr_genes$age$cor_10_Me__14_Premet$heatmap
)

save_plot(
  file = "graphics/pfla_blan_hicor.pdf",
  base_height = 15,
  base_width = 10,
  plot_grid(p1,p2,p3,ncol = 1)
)


pdf("graphics/pfla_blan_Mecthnikoff_Premetamorphic.pdf",
    width = 10, height = 5)
plot_grid(
  PFLA_BLAN_COMPARISON$high_corr_genes$pairwise_data$hicor_topgenes$cor_10_Me__14_Premet$plot_topgenes,
  PFLA_BLAN_COMPARISON$high_corr_genes$GOs$GOplot$cor_10_Me__14_Premet,
  ncol = 2, align = "v", axis = "tb", rel_widths = c(1,2)
  )
dev.off()


pfsp_cors_oma <- grid.grabExpr(plot_cors(PFLA_SPUR_COMPARISON$pairwise_correlations))
pfsp_cors_rbh <- grid.grabExpr(plot_cors(pf_sp_rbh_cors_allgen))
pfbl_cors_oma <- grid.grabExpr(plot_cors(PFLA_BLAN_COMPARISON$pairwise_correlations))
pfbl_cors_rbh <- grid.grabExpr(plot_cors(pf_bl_rbh_cors_allgen))

save_plot(
  file = "graphics/suppl_cors_oma_rbh.pdf",
  base_height = 10,
  base_width = 8,
  plot_grid(
    pfsp_cors_oma,
    pfsp_cors_rbh,
    pfbl_cors_oma,
    pfbl_cors_rbh,
    ncol = 1
  )
)

ggarrange(
  pfla_tfs_ngenes_plot,
  spur_tfs_ngenes_plot,
  blan_tfs_ngenes_plot,
  pfla_tfs_expgenes_plot,
  spur_tfs_expgenes_plot,
  blan_tfs_expgenes_plot,
  pfla_tf_EXPNGEN_plot,
  spur_tf_EXPNGEN_plot,
  blan_tf_EXPNGEN_plot,
  ncol=3,
  nrow=3,
  common.legend=T
) %>%
ggexport(filename = "graphics/suppl_barplot_TFs_3species.pdf")

```