---
title: "Ptychodera Cisreg Development: ATAC-seq Mapping Quality Analysis"
author: "Alberto Perez-Posada"
date: "3/31/2023"
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}

dir <- '/home/ska/aperpos/projects/ptychodera_cisreg_development/'
fcha <- function(){ gsub("-","",Sys.Date()) }

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = dir)
options(scipen=999)
Sys.setenv(VROOM_CONNECTION_SIZE=5000072)

```


## About

This markdown generates the panels showcasing the mapping statistics of the chromatin accessibility data.

## Load libraries

```{r libraries, message=FALSE, warning=FALSE}
library(DESeq2)
library(limma)
library(ComplexHeatmap)
library(circlize)
library(ggplot2)
```

## Load functions

```{r load_functions, message = FALSE}
source("code/r_code/r_functions/sourcefolder.R")

sourceFolder(
  "code/r_code/r_functions",
  recursive = TRUE
  )

sourceFolder(
  "code/r_code/r_general",
  recursive = TRUE
  )
```

## ATAC seq quality analysis

```{r}
# load table
pfla_atac_qual <- read.delim2(
  "./outputs/reports_atac/atac_quality.tsv",
  header=T
)
```

### Bowtie2 Statistics

Below are the stats of the bowtie2 mapping.

```{r}
pfla_atac_bowtie2 <- pfla_atac_qual[,1:4]

pfla_atac_bowtie2
```

To plot this we can relativise every metrc by the total number of reads per sample. This way we can see the percentages.

```{r, echo = FALSE}
pfla_atac_bowtie2$pct_unmapped <- apply(
  pfla_atac_bowtie2[,2:4],
  1,
  function(x){x[1]/sum(x)}
)
pfla_atac_bowtie2$pct_singlemap <- apply(
  pfla_atac_bowtie2[,2:4],
  1,
  function(x){x[2]/sum(x)}
)
pfla_atac_bowtie2$pct_multimap <- apply(
  pfla_atac_bowtie2[,2:4],
  1,
  function(x){x[3]/sum(x)}
)
```

And the plot itself:

```{r}
barplot(
  as.matrix(t(pfla_atac_bowtie2[,5:7])),
  names.arg = pfla_atac_bowtie2$sample,
  main = "Ptychodera flava ATAC-seq mapping quality",
  xlab = "Sample",
  ylab = "% Reads mapped",
  legend.text = c("unmapped","single-mapping","multi-mapping"),
  args.legend = list(x = "topright",
                     inset = c(- 0.15, 0)),
  col = c("lightgray","royalblue","lightblue"),
  xlim = c(0,11)
)
```

We can see that these are not stellar golden-standard level stats, but it is what we have.

### IDR statistics

A subset of the stats in the original table are for IDR.

```{r}
pfla_atac_idr <- pfla_atac_qual[c(1,3,5,7),c(1,5:17)]

pfla_atac_idr
```

To plot these:

```{r, echo = FALSE}
barplot(
  t(pfla_atac_idr[,2:12])/1000,
  beside = T,
  names.arg = rep(colnames(pfla_atac_idr)[2:12], 4),
  las = 2,
  col = c(rep(brewer.pal(4,"Spectral")[1],11), rep(brewer.pal(4,"Spectral")[2],11), rep(brewer.pal(4,"Spectral")[3],11), rep(brewer.pal(4,"Spectral")[4],11)),
  main = "IDR stats",
  legend.text = c("EB","EG","MG","LG"),
  args.legend = list(x = "topright", inset = c(- 0.1, 0), fill = brewer.pal(4,"Spectral")),
  ylab = "# peaks (x10E3)"
)
```


```{r, echo = FALSE}
barplot(
  height = as.numeric(t(pfla_atac_idr[,13:14])),
  beside = T,
  names.arg = rep(colnames(pfla_atac_idr)[13:14], 4),
  las = 2,
  col = sort(rep(brewer.pal(4,"Spectral"),2)),
  main = "IDR stats",
  legend.text = c("EB","EG","MG","LG"),
  args.legend = list(x = "topright", inset = c(- 0.1, 0), fill = sort(brewer.pal(4,"Spectral"))),
  ylab = "Index"
)
```
