---
title: "Ptychodera Cisreg Development: Differential Gene Expression Analysis"
author: "Alberto Perez-Posada"
date: "3/24/2023"
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}

dir <- '/home/ska/aperpos/projects/ptychodera_cisreg_development/'
fcha <- function(){ gsub("-","",Sys.Date()) }

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = dir)
options(scipen=999)
Sys.setenv(VROOM_CONNECTION_SIZE=5000072)

```


## About



## Load libraries

```{r libraries, message=FALSE, warning=FALSE}
library(DESeq2)
library(ComplexHeatmap)
library(circlize)
library(topGO)
```

## Load functions

```{r load_functions, message=FALSE, warning=FALSE}
#from @Rich Scriven https://stackoverflow.com/questions/26045478/source-r-scripts-in-different-folders
sourceFolder <- function(folder, recursive = FALSE, ...) 
{ 
    files <- list.files(folder, pattern = "[.][rR]$", 
                        full.names = TRUE, recursive = recursive)
    if (!length(files))
        stop(simpleError(sprintf('No R files in folder "%s"', folder)))
    src <- invisible(lapply(files, source, ...))
    message(sprintf('%s files sourced from folder "%s"', length(src), folder))
}

sourceFolder(
  "code/r_code/r_functions",
  recursive = TRUE
  )

sourceFolder(
  "code/r_code/r_general",
  recursive = TRUE
  )
```


## Data Load

We will load the normalised counts for DESeq2:

```{r load data}
load("outputs/rda/normalised_counts.rda")
```

## Correlation plots

We can explore the structure of our data by doing correlation as well as dimensionality reduction.

```{r spearman}
rna_counts_sp <- 
  cor(
    sapply(
      levels(condition_x),
      rowMeans_by_repl,
      tbl = pfla_rna_counts
    )
  )
```

To summarise our data, we will calculate the arithmetic mean across columns that are replicates from the same developmental stage using a custom function. Thus we end up with:

```{r heatmap sp}
pf_sp_hm <-
  Heatmap(
    rna_counts_sp,
    name="Spearman",
    top_annotation=devstages_ha_columns(),
    left_annotation=devstages_ha_rows(),
    cluster_rows = F,
    cluster_columns = F,
    show_row_names=F,
    show_column_names=F,
    col=colorRampPalette(rev(brewer.pal(7,"RdYlBu")))(100)
  )

draw(pf_sp_hm)
```

## DESeq2 analysis

We will import the RUV-normalised counts to DESeq2 following the design matrix we generated before.

```{r deseqdataset}
dds <- 
  DESeqDataSetFromMatrix(
    countData = pfla_rna_counts,
    colData = pData(set2),
    design = ~ W_1 + condition_x
    )

dds <- estimateSizeFactors(dds)

nc <- counts(dds, normalized=TRUE)

filter <- rowSums(nc >= 10) >= 2

dds <- dds[filter,]
```

Estimation of factor sizes and the dispersion

```{r estimations}
dds <- estimateSizeFactors(dds)
dds <- estimateDispersions(dds)
```

The likelihood-ratio test is the proper one for time-series analysis.

```{r LRT test}
dds <- 
  nbinomLRT(
    dds,
    reduced=as.formula("~ W_1"),
    maxit=500
    )
```

And the results:

```{r results}
res <- results(dds)

round(data.frame(res)[1:10,],3)
```

We will apply a cutoff to keep genes with a p-value < 0.05 (likelihood ratio test) as significant.

```{r signif}
cutoff=0.05

res_signif <- 
  subset(res, padj < cutoff )

head(apply(res_signif,2,round,digits=3))
```

To proceed with data visualisation and clustering analysis, we will use the variance-stabilised transform of this data.

```{r vst}
# Variance-stabilised transformation
vsd <- vst(dds)

# Assay with all genes
vsd_allgen <- 
  as.data.frame(
    assay(vsd)
    )

# Assay with only significant genes
vsd_signif <- 
  as.data.frame(
    assay(
      vsd[rownames(vsd) %in% rownames(res_signif),]
      )
    )

```

To summarise our data, we will calculate the arithmetic mean across columns that are replicates from the same developmental stage using a custom function. Thus we end up with:

```{r rowmeans_byrepl}
pfla_rna_all <-
    as.data.frame(
    sapply(
      levels(condition_x),
      rowMeans_by_repl,
      tbl = vsd_allgen # all genes
      )
  )

rownames(pfla_rna_all) <- rownames(vsd_allgen)

pfla_rna_dev <-
  data.frame(
    sapply(
      levels(condition_x),
      rowMeans_by_repl,
      tbl = vsd_signif # significantly regulated throughout develompent
      )
  )

rownames(pfla_rna_dev) <- rownames(vsd_signif)

pfla_rna_dev[1:5,1:5]

```

We can cluster our data to detect patterns of gene expression during development.

```{r hcluster}

pfla_rna_dev_hc <- 
  hclust(
    dist(t(scale(t(pfla_rna_dev)))),
    method = "ward.D2"
  )

```

We pick a threshold to cut around distance "50".

```{r plot hcluster}

plot(pfla_rna_dev_hc,cex=0.3)

cutoff = 50
abline(cutoff, 0, col="red")

```

And we proceed to binning the genes in clusters:

```{r cID}
pfla_rna_dev$clusterID <-
  cutree(pfla_rna_dev_hc, h = cutoff)

unique(pfla_rna_dev$clusterID)
```

We manually rearranged these clusters and provided new IDs.

```{r}

hclu_names <- 
  read.table(
    "assets/rna_dev_hcluster_name_equiv.tsv",
    header = F, 
    col.names = c("clusterID","cID"),
    colClasses = c("numeric","character")
  )

pfla_rna_dev$cID <-
  hclu_names$cID[
    match(
      pfla_rna_dev$clusterID,
      as.integer(hclu_names$clusterID)
      )
    ]

#remove old numbering
pfla_rna_dev$clusterID <- NULL
```

We will subsample this gigantic dataset to retrieve a few genes per cluster in order to see their dynamics during development.

```{r}
set.seed(2)

pfla_rna_dev_viz <- 
  pfla_rna_dev[
    sample(
      nrow(pfla_rna_dev),
      200
    ),
  ]
```

Lorem Ipsum

```{r heatmap, fig.width = 4, fig.height = 7.5, warning=FALSE}
pf_hm <- 
  Heatmap(
    t(scale(t(pfla_rna_dev_viz[,1:16]))),
    name = "gene exp",
    top_annotation=devstages_ha_columns(),
    row_split=pfla_rna_dev_viz$cID,
    show_row_names = F,
    show_column_names=F,
    cluster_rows=F,
    cluster_columns=F,
    column_names_side="top",
    row_title = NULL,
    col=app_pal1,
    use_raster = F
  )

draw(pf_hm)
```

A tidier visualisation of the clusters that we will use throughout the manuscript. This is achieved averaging the expression profile of all genes of each clusters.

```{r}

pfla_rna_dev_avg <-
  data.frame(
    aggregate(
      t(scale(t(pfla_rna_dev[,1:16]))),
      by = list(cluster = pfla_rna_dev$cID),
      FUN = mean
    )[-1],
    row.names = sort(unique(pfla_rna_dev$cID))
  )

```

Lorem Ipsum

```{r heatmap_avg, fig.width = 4, fig.height = 5}
pf_avg_hm <-
  Heatmap(
    name="avg exp\nz-score",
    as.matrix(quantile_normalisation(pfla_rna_dev_avg)),
    show_row_names = TRUE,
    row_names_side = "left",
    show_column_names = F,
    cluster_rows=F,
    cluster_columns=F,
    top_annotation=devstages_ha_columns(),
    column_names_side="top",
    row_title = NULL,
    col=cividis(7)
  )

draw(pf_avg_hm)
```

## Gene Ontology Analysis of stage-specific gene clusters

After this, we do Gene Ontology using a custom wrapper function:

```{r, warning = FALSE, message = FALSE}
library(topGO)

#gene universe
gene_universe <- allgenes

# gene-GO mappings
pfla_id_GO <-
  readMappings(
    "outputs/functional_annotation/go_blast2go/GO_annotation.txt"
  )

#list of genes of interest
pfla_clusters_list <- list()

for (i in unique(pfla_rna_dev$cID)) {
  pfla_clusters_list[[i]] <- 
    rownames(pfla_rna_dev)[pfla_rna_dev$cID == i]
}

names(pfla_clusters_list) <- unique(pfla_rna_dev$cID)

pfla_GOs <- getGOs(
  genelist = pfla_clusters_list,
  gene_universe = gene_universe,
  gene2GO = pfla_id_GO
)

```

## Saving the data

Saving the data for follow-up analyses

```{r}
save(
  #counts
  pfla_rna_counts,
  condition_x,
  #DESeq2 results
  dds,
  res,
  res_signif,
  #VST data
  vsd_allgen,
  vsd_signif,
  pfla_rna_all,
  #stage-specific gene expression dataset
  pfla_rna_dev,
  pfla_rna_dev_viz,
  pfla_rna_dev_avg,
  #Heatmaps
  pf_sp_hm,
  pf_hm,
  pf_avg_hm,
  #GO analysis
  pfla_GOs,
  #Output file
  file = "outputs/rda/deseq2.rda"
)

```

A smaller version of this output with just the counts and the stage-specific clusters:

```{r}
save(
  allgenes,
  # counts
  pfla_rna_counts,
  condition_x,
  #VST data
  pfla_rna_all,
  #stage-specific gene expression dataset
  pfla_rna_dev,
  pfla_rna_dev_avg,
  #Heatmaps
  pf_hm,
  pf_avg_hm,
  #GO analysis
  pfla_GOs,
  #Output file
  file = "outputs/rda/stage_specific_clusters.rda"
)

```
